---
title: "Seurat workflows for single and merged samples"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    code_folding: show
---

```{r document-setup, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(
  eval = FALSE, echo = TRUE, cache = FALSE, tidy = FALSE, error = FALSE
)
```

# Using this document

This R Markdown document explains two Seurat 5 workflows: an independent
single-sample analysis and a pooled analysis of three samples. Each workflow
loads filtered raw UMI counts, records initial cell quality, normalises counts,
calculates PCA and UMAP, and saves a Seurat object. Choose the workflow that
matches the inputs. The single sample is separate from the three-sample merge.

**Status:** this is code with explanations. Preparing this document has not run
either analysis, and no cell counts, plots or biological findings are reported.
The R code in each workflow reproduces its matching script in `scripts/`.

Open this file in RStudio to read the explanation alongside the code. With
`rmarkdown`, `knitr` and Pandoc available, **Knit** creates an HTML guide showing
the code without running the analysis. Every analysis chunk explicitly uses
`eval=FALSE`; the hidden first chunk only sets document options. This follows
the [knitr chunk-option documentation](https://yihui.org/knitr/options/).

For an intentional analysis, use a dedicated interactive RStudio session and
set its working directory to the repository root. Preserve any other active
analysis session. Run the chosen workflow's chunks in order, inspecting the
outputs before continuing. RStudio's interactive Run controls execute code;
`eval=FALSE` controls knitting, not those manual Run controls. Keep the other
workflow unexecuted. Both workflows retain their RStudio checks and end at
`# STOP HERE` after `saveRDS()`.

## Inputs and packages

The analysis needs `Seurat` version 5 or later, `SeuratObject`, `dplyr`, `ggplot2`
and `patchwork`. Package checks report missing software without installing it.
Knitting the guide alone does not require Seurat or input matrices.

Use the following directories relative to the repository root:

```text
data/
  single_sample/filtered_feature_bc_matrix/
  sample_1/filtered_feature_bc_matrix/
  sample_2/filtered_feature_bc_matrix/
  sample_3/filtered_feature_bc_matrix/
```

Each matrix directory contains matching `matrix.mtx.gz`, `features.tsv.gz` and
`barcodes.tsv.gz` files. The single-sample workflow also accepts uncompressed
versions. These must be raw UMI counts, not previously normalised values. The
sample names are editable placeholders; no organism or biological group is
assumed. The [README](README.md) explains how the terminal Cell Ranger script
produces these inputs.

## Parameters to review

| Step | Setting in the code | Meaning |
|---|---|---|
| Object creation | `min.cells = 3` | Retain features detected in at least three cells within an input sample. |
| Object creation | `min.features = 200` | Initially retain cells with at least 200 detected features. |
| Subsequent cell filter | `nFeature_RNA > 200` | Retain cells with strictly more than 200 detected features. |
| Normalisation | `LogNormalize`, scale factor 10,000 | Adjust counts for cell totals, scale and log-transform. |
| Variable features | `vst`, 2,000 requested | Identify features with high variability for PCA. |
| Scaling and PCA | Scale all retained genes; use variable features for PCA | Separate the scaling scope from the PCA input. |
| UMAP | PCs 1–10; random seed 42 | Calculate an exploratory embedding from the selected PCs. |

These settings are examples to review against the data. No mitochondrial or
upper-count filter is included. Inspect quality distributions before deciding
on further filtering. Enough cells and variable features must remain for PCA
and UMAP, and scaling all genes can require substantial memory. A fixed seed
helps repeatability within a recorded software environment; software versions
still matter. The [Seurat introductory workflow](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html)
provides method background.

# Single-sample workflow

This section corresponds to [single_sample_02_seurat.R](scripts/single_sample_02_seurat.R).
It uses only `data/single_sample/filtered_feature_bc_matrix/` and creates a new
folder under `results/single_sample/` for each run.

## Check the RStudio environment

Check the session and required packages before reading data. Seurat 5 assay
behaviour is selected explicitly, and the random seed is set to 42. If a
check stops the workflow, resolve the reported problem before proceeding.

```{r single-environment, eval=FALSE}
# Single-sample Seurat analysis
# Open in RStudio, set the working directory to the repository root,
# and review each section before running it.
# Each run writes to a new results folder.

# 1. Check the RStudio session and installed packages -------------------------
if (!interactive() || Sys.getenv("RSTUDIO") != "1") {
  stop("Open this script in RStudio before running it.")
}
required_packages <- c("Seurat", "dplyr", "ggplot2", "patchwork")
missing_packages <- required_packages[!vapply(
  required_packages, requireNamespace, logical(1), quietly = TRUE
)]
if (length(missing_packages) > 0) {
  stop("Missing packages: ", paste(missing_packages, collapse = ", "),
       ". Make these packages available in RStudio before continuing.")
}
library(Seurat)
library(dplyr)
library(ggplot2)
library(patchwork)
if (packageVersion("Seurat") < "5.0.0") {
  stop("This script requires Seurat 5 or later.")
}
options(Seurat.object.assay.version = "v5")
set.seed(42)

```

## Select inputs and prepare output records

Read the input location from the working directory and `sample_id`. Verify
the matrix files, then create a timestamped folder and record package
versions and run settings. The `save_plot()` helper displays each plot in
RStudio and writes a PNG. Inspect the printed paths before continuing.

```{r single-inputs, eval=FALSE}
# 2. Select the input and create a new output folder -----------------
project_root <- getwd()
sample_id <- "single_sample"
matrix_dir <- file.path(project_root, "data", sample_id,
                        "filtered_feature_bc_matrix")
if (!dir.exists(matrix_dir)) {
  stop("Cannot find the input matrix: ", matrix_dir)
}
matrix_files <- file.path(matrix_dir, c("matrix.mtx", "features.tsv", "barcodes.tsv"))
if (any(!file.exists(matrix_files) & !file.exists(paste0(matrix_files, ".gz")))) {
  stop("The matrix folder is missing a matrix, feature, or barcode file.")
}

run_stamp <- format(Sys.time(), "%Y%m%d_%H%M%S")
run_dir <- file.path(project_root, "results", "single_sample",
                     paste0("run_", run_stamp))
if (dir.exists(run_dir)) {
  stop("This output folder already exists. Run this section again for a new timestamp.")
}
if (!dir.create(run_dir, recursive = TRUE)) stop("Cannot create output folder: ", run_dir)
message("Input matrix: ", matrix_dir)
message("This run's outputs: ", run_dir)

package_versions <- data.frame(
  package = required_packages,
  version = vapply(required_packages, function(x) as.character(packageVersion(x)), "")
)
utils::write.csv(package_versions, file.path(run_dir, "package_versions.csv"), row.names = FALSE)
writeLines(c(
  paste("Started:", format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")),
  paste("R:", R.version.string),
  paste("RStudio session:", Sys.getenv("RSTUDIO")),
  paste("Input:", matrix_dir),
  "Initial settings: min.cells=3; min.features=200; nFeature_RNA > 200.",
  "LogNormalize scale.factor=10000; vst nfeatures=2000; scale all retained genes.",
  "PCA uses variable features; UMAP uses PCs 1:10; random seed=42.",
  "No mitochondrial filter, clustering, or cell annotation is applied."
), file.path(run_dir, "run_details.txt"))

# Print plots explicitly so RStudio Source displays them, and save a PNG copy.
save_plot <- function(plot, filename, width = 8, height = 5) {
  print(plot)
  ggplot2::ggsave(file.path(run_dir, filename), plot = plot,
                  width = width, height = height, units = "in", dpi = 300)
}

```

## Read counts and create a Seurat object

`Read10X()` loads the count matrix. If it returns several measurement
types, select Gene Expression. Record input dimensions, create the object
with the initial feature and cell filters, and retain the sample ID.
The printed object describes cells and features actually retained.

```{r single-object, eval=FALSE}
# 3. Read the Cell Ranger matrix and create the Seurat object -----------------
counts <- Seurat::Read10X(data.dir = matrix_dir)
if (is.list(counts)) {
  if (!"Gene Expression" %in% names(counts)) stop("Gene Expression matrix not found.")
  counts <- counts[["Gene Expression"]]
}
# Check the supplied matrix without assuming dataset-specific dimensions.
if (nrow(counts) == 0L || ncol(counts) == 0L) {
  stop("The input matrix has no features or barcodes.")
}
cell_counts <- data.frame(
  stage = "Cell Ranger filtered matrix",
  features = nrow(counts), cells = ncol(counts)
)
seurat_object <- Seurat::CreateSeuratObject(
  counts = counts, project = sample_id, min.cells = 3, min.features = 200
)
seurat_object$sample_id <- sample_id
cell_counts <- rbind(cell_counts, data.frame(
  stage = "CreateSeuratObject min.cells 3 and min.features 200",
  features = nrow(seurat_object), cells = ncol(seurat_object)
))
print(seurat_object)

```

## Inspect quality and filter cells

Violin plots show the distribution of detected genes and total UMI counts;
the scatter plot shows their relationship. Inspect these before applying
the strict `nFeature_RNA > 200` filter. The stage-count CSV records the
effect of filtering, and the metadata CSV retains per-cell measurements.
These plots alone do not establish a complete quality-control decision.

```{r single-qc, eval=FALSE}
# 4. Inspect cell quality and apply the initial feature filter --------
p_qc_violin <- Seurat::VlnPlot(
  seurat_object, features = c("nFeature_RNA", "nCount_RNA"), ncol = 2
)
save_plot(p_qc_violin, "qc_violin_before_filter.png")
p_qc_scatter <- Seurat::FeatureScatter(
  seurat_object, feature1 = "nCount_RNA", feature2 = "nFeature_RNA"
)
save_plot(p_qc_scatter, "qc_scatter_before_filter.png")

# This example retains cells with more than 200 detected genes.
# Select additional thresholds only after inspecting the data.
# No mitochondrial or upper-count filter is applied.
seurat_object <- subset(seurat_object, subset = nFeature_RNA > 200)
cell_counts <- rbind(cell_counts, data.frame(
  stage = "After nFeature_RNA greater than 200",
  features = nrow(seurat_object), cells = ncol(seurat_object)
))
print(cell_counts)
utils::write.csv(cell_counts, file.path(run_dir, "cell_counts_by_stage.csv"), row.names = FALSE)
utils::write.csv(seurat_object[[]], file.path(run_dir, "cell_metadata_after_filter.csv"))

```

## Normalise and find variable features

Log-normalise counts and request 2,000 variable features using `vst`.
The two plots show feature variability, with the top ten labelled in the
second plot. Their names are also written to a CSV. These are variable
features for dimensional reduction; no cell-type markers are identified.

```{r single-normalise, eval=FALSE}
# 5. Normalise counts and identify 2,000 variable features --------------------
seurat_object <- Seurat::NormalizeData(
  seurat_object, normalization.method = "LogNormalize", scale.factor = 1e4
)
seurat_object <- Seurat::FindVariableFeatures(
  seurat_object, selection.method = "vst", nfeatures = 2000
)
top_vf <- head(Seurat::VariableFeatures(seurat_object), 10)
print(top_vf)
utils::write.csv(data.frame(gene = top_vf),
                 file.path(run_dir, "top10_variable_features.csv"), row.names = FALSE)
p_variable <- Seurat::VariableFeaturePlot(seurat_object)
p_variable_labels <- Seurat::LabelPoints(
  plot = p_variable, points = top_vf, repel = TRUE, xnudge = 0, ynudge = 0
)
save_plot(p_variable, "variable_features.png")
save_plot(p_variable_labels, "variable_features_top10.png")

```

## Scale genes and examine PCA

Scale all retained genes, then calculate PCA from the selected variable
features. Save the leading feature loadings, loading plots, cell embedding
and a PC 1 heatmap. Inspect which genes drive the leading components and
whether technical variation may contribute before using PCs 1–10.

```{r single-pca, eval=FALSE}
# 6. Scale all retained genes and perform PCA on variable features ------------
# Scaling every retained gene uses more memory than
# scaling only the variable genes. PCA still uses the selected variable genes.
all.genes <- rownames(seurat_object)
seurat_object <- Seurat::ScaleData(seurat_object, features = all.genes)
seurat_object <- Seurat::RunPCA(
  seurat_object, features = Seurat::VariableFeatures(object = seurat_object),
  seed.use = 42
)
print(seurat_object[["pca"]], dims = 1:5, nfeatures = 5)
writeLines(capture.output(print(seurat_object[["pca"]], dims = 1:5, nfeatures = 5)),
           file.path(run_dir, "pca_top_features.txt"))

p_pca_loadings <- Seurat::VizDimLoadings(seurat_object, dims = 1:2, reduction = "pca")
save_plot(p_pca_loadings, "pca_loadings_dims1_2.png", width = 10, height = 6)
p_pca <- Seurat::DimPlot(seurat_object, reduction = "pca")
save_plot(p_pca, "pca_cells.png")
# fast=FALSE returns a plot that can be printed and saved with ggsave.
# Display PC 1 with 500 cells and balanced feature selection.
p_pca_heatmap <- Seurat::DimHeatmap(
  seurat_object, dims = 1, cells = 500, balanced = TRUE, fast = FALSE
)
save_plot(p_pca_heatmap, "pca_heatmap_dim1.png", width = 10, height = 7)

```

## Calculate an exploratory UMAP

Use PCs 1–10 to calculate UMAP and save the cell embedding. Nearby points
represent similarity in this reduced representation. The plot does not
assign clusters or cell types, and visual separation is not a statistical
test of a biological difference.

```{r single-umap, eval=FALSE}
# 7. Calculate and display UMAP using PCs 1-10 -------------------
# The seed improves repeatability within the recorded software environment.
# This is an exploratory embedding; no clusters or cell types are assigned.
seurat_object <- Seurat::RunUMAP(seurat_object, dims = 1:10, seed.use = 42)
p_umap <- Seurat::DimPlot(seurat_object, reduction = "umap")
save_plot(p_umap, "umap_cells.png")

```

## Record the session and save the object

After the preceding steps succeed, save the actual session information
and the Seurat object as `single_sample_seurat_object.rds` in the run
folder. The RDS preserves the object for later authorised work. This is
the final analysis step in the single-sample workflow.

```{r single-save, eval=FALSE}
# 8. Record the session and save the object
# Run this final section only after the earlier sections complete successfully.
writeLines(capture.output(sessionInfo()), file.path(run_dir, "sessionInfo.txt"))
rds_path <- file.path(run_dir, "single_sample_seurat_object.rds")
message("Saving the Seurat object to: ", rds_path)
saveRDS(seurat_object, file = rds_path)
# STOP HERE
```

# Three-sample merged workflow

This section corresponds to [merged_02_seurat.R](scripts/merged_02_seurat.R).
It uses `sample_1`, `sample_2` and `sample_3` only. Each input is filtered at
object creation before the objects are combined for pooled processing.
The independent `single_sample` input is not part of this merge.

## Check the environment and configure three inputs

Check RStudio and packages, then define the three sample IDs, labels and
matrix paths in `sample_info`. Verify the compressed matrix files and
create a new folder under `results/merged/`. The input table and initial
session information document the configuration actually used.

```{r merged-setup, eval=FALSE}
# Three-sample Seurat analysis
# Open in RStudio, set the working directory to the repository root,
# and review each section before running it.
# Inputs are filtered feature-barcode matrices containing raw UMI counts.
# merge() combines the samples without batch correction or integration.

# Setup -------------------------------------------------------------------
if (!interactive() || Sys.getenv("RSTUDIO") != "1") {
  stop("Open this script in RStudio and run it there, not with Rscript.")
}
required_packages <- c("Seurat", "SeuratObject", "dplyr", "ggplot2", "patchwork")
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages) > 0L) {
  stop("Missing packages: ", paste(missing_packages, collapse = ", "),
       ". Make these packages available in RStudio before continuing.")
}
library(dplyr)
library(ggplot2)
library(Seurat)
library(patchwork)
if (packageVersion("Seurat") < "5.0.0") {
  stop("This merged workflow requires Seurat 5 or later.")
}
options(Seurat.object.assay.version = "v5")
set.seed(42)

project_dir <- getwd()
sample_info <- data.frame(
  sample_id = c("sample_1", "sample_2", "sample_3"),
  sample_label = c("sample_1", "sample_2", "sample_3"),
  stringsAsFactors = FALSE
)
sample_info$matrix_dir <- file.path(
  project_dir, "data", sample_info$sample_id, "filtered_feature_bc_matrix"
)
required_files <- c("matrix.mtx.gz", "features.tsv.gz", "barcodes.tsv.gz")
for (matrix_dir in sample_info$matrix_dir) {
  if (!all(file.exists(file.path(matrix_dir, required_files)))) {
    stop("A required matrix file is missing from: ", matrix_dir)
  }
}

# Each run has its own output folder so previous results are preserved.
run_id <- paste0("merged_", format(Sys.time(), "%Y%m%d_%H%M%S"))
output_dir <- file.path(project_dir, "results", "merged", run_id)
if (dir.exists(output_dir)) stop("Output folder already exists: ", output_dir)
if (!dir.create(output_dir, recursive = TRUE)) stop("Cannot create: ", output_dir)
write.csv(sample_info, file.path(output_dir, "sample_inputs.csv"), row.names = FALSE)
capture.output(sessionInfo(), file = file.path(output_dir, "sessionInfo_start.txt"))
message("Results will be saved in: ", output_dir)

```

## Create an object for each sample

Read each Gene Expression matrix and apply `min.cells = 3` and
`min.features = 200` separately within that sample. Store both sample ID
and label in cell metadata. Record input and retained cell counts for
each sample so unequal losses can be inspected later.

```{r merged-objects, eval=FALSE}
# Read matrices and create one object per sample ---------------------------
# Initial filters: genes detected in >=3 cells within each sample, and cells
# with >=200 detected genes at object creation. We record those removals.
sample_objects <- setNames(vector("list", nrow(sample_info)), sample_info$sample_id)
cell_counts <- data.frame(sample_id = sample_info$sample_id,
                          cellranger_barcodes = NA_integer_,
                          after_creation = NA_integer_)
for (i in seq_len(nrow(sample_info))) {
  counts <- Read10X(data.dir = sample_info$matrix_dir[i])
  if (is.list(counts)) {
    if (!"Gene Expression" %in% names(counts)) stop("Gene Expression counts missing.")
    counts <- counts[["Gene Expression"]]
  }
  cell_counts$cellranger_barcodes[i] <- ncol(counts)
  sample_object <- CreateSeuratObject(
    counts = counts, project = sample_info$sample_id[i],
    min.cells = 3, min.features = 200
  )
  if (ncol(sample_object) == 0L) stop("No cells retained for ", sample_info$sample_id[i])
  sample_object$sample_id <- sample_info$sample_id[i]
  sample_object$sample_label <- sample_info$sample_label[i]
  cell_counts$after_creation[i] <- ncol(sample_object)
  sample_objects[[i]] <- sample_object
}

```

## Merge samples and join count layers

Merge the three objects with sample prefixes on barcodes, then check
that barcodes are unique and the total cell count is preserved.
`JoinLayers()` joins the Seurat 5 count layers for pooled variable-feature
selection. The layer record documents this change. Sample labels remain
available for plots. This operation does not correct batch effects or
establish comparability of the experiments. See the
[merge reference](https://satijalab.github.io/seurat-object/reference/merge.Seurat.html)
and [JoinLayers reference](https://satijalab.github.io/seurat-object/reference/SplitLayers.html).

```{r merged-join, eval=FALSE}
# Merge all three samples before the common analysis ----------------------
# Prefix barcodes to keep them unique and preserve each cell's sample labels.
seurat_merged <- merge(
  x = sample_objects[["sample_1"]],
  y = list(sample_objects[["sample_2"]], sample_objects[["sample_3"]]),
  add.cell.ids = sample_info$sample_id,
  project = "merged_samples", merge.data = FALSE
)
stopifnot(ncol(seurat_merged) == sum(cell_counts$after_creation),
          !anyDuplicated(colnames(seurat_merged)))

# Seurat 5 preserves separate count layers when merging. Join counts into one
# layer so variable-feature selection uses the pooled dataset rather than a
# consensus of per-sample variable features.
# API: https://satijalab.github.io/seurat-object/reference/SplitLayers.html
layers_before <- SeuratObject::Layers(seurat_merged[["RNA"]])
seurat_merged <- SeuratObject::JoinLayers(seurat_merged, assay = "RNA")
layers_after <- SeuratObject::Layers(seurat_merged[["RNA"]])
stopifnot(identical(layers_after, "counts"))
capture.output(list(before_join = layers_before, after_join = layers_after),
               file = file.path(output_dir, "merged_layers.txt"))
seurat_merged$sample_label <- factor(
  seurat_merged$sample_label, levels = c("sample_1", "sample_2", "sample_3")
)
Idents(seurat_merged) <- "sample_label"
rm(counts, sample_object, sample_objects)

```

## Inspect quality by sample and filter cells

Plot detected features and total counts by sample, then apply the strict
`nFeature_RNA > 200` filter to the merged object. Save metadata before and
after filtering, and record retained cells per sample. If any sample has
no cells remaining, the workflow stops for review. Interpret differences
in retention alongside the input quality and experimental design.

```{r merged-qc, eval=FALSE}
# QC plots and the initial feature filter -----------------------------------
qc_violin <- VlnPlot(
  seurat_merged, features = c("nFeature_RNA", "nCount_RNA"),
  group.by = "sample_label", ncol = 2
)
print(qc_violin)
ggsave(file.path(output_dir, "qc_violin_before_filter.png"), qc_violin,
       width = 10, height = 5, dpi = 300)
qc_scatter <- FeatureScatter(
  seurat_merged, feature1 = "nCount_RNA", feature2 = "nFeature_RNA",
  group.by = "sample_label"
)
print(qc_scatter)
ggsave(file.path(output_dir, "qc_scatter_before_filter.png"), qc_scatter,
       width = 7, height = 5, dpi = 300)
write.csv(seurat_merged[[]], file.path(output_dir, "cell_metadata_before_filter.csv"))

# This example retains cells with more than 200 detected genes.
# No mitochondrial or upper-count threshold is applied.
# Choose suitable thresholds after inspecting the data.
seurat_merged <- subset(seurat_merged, subset = nFeature_RNA > 200)
cell_counts$after_feature_filter <- as.integer(table(factor(
  seurat_merged$sample_id, levels = sample_info$sample_id
)))
print(cell_counts)
write.csv(cell_counts, file.path(output_dir, "cell_counts_by_stage.csv"), row.names = FALSE)
write.csv(seurat_merged[[]], file.path(output_dir, "cell_metadata_after_filter.csv"))
if (any(cell_counts$after_feature_filter == 0L)) {
  stop("At least one sample has no cells after filtering. Review QC before continuing.")
}

```

## Normalise pooled counts and find variable genes

Apply LogNormalize to the merged object and select variable genes from
the joined assay. Save the selected genes and a variability plot with
the top ten labelled. Review whether selected features may reflect
technical differences as well as biology; this step is not integration.

```{r merged-normalise, eval=FALSE}
# Normalise and identify variable genes -----------------------------------
seurat_merged <- NormalizeData(
  seurat_merged, normalization.method = "LogNormalize", scale.factor = 1e4
)
seurat_merged <- FindVariableFeatures(
  seurat_merged, selection.method = "vst", nfeatures = 2000
)
top_vf <- head(VariableFeatures(seurat_merged), 10)
print(top_vf)
write.csv(data.frame(gene = VariableFeatures(seurat_merged)),
          file.path(output_dir, "variable_genes.csv"), row.names = FALSE)
plot1 <- VariableFeaturePlot(seurat_merged)
plot2 <- LabelPoints(plot1, points = top_vf, repel = TRUE, xnudge = 0, ynudge = 0)
print(plot1)
print(plot2)
ggsave(file.path(output_dir, "variable_genes_top10.png"), plot2,
       width = 8, height = 6, dpi = 300)

```

## Examine PCA and sample relationships

Scale all retained genes and calculate PCA from variable features. Save
loadings, a PCA plot coloured by sample and a PC 1 heatmap. Compare sample
distributions and inspect the genes associated with the components.
Sample separation can have biological or technical explanations.

```{r merged-pca, eval=FALSE}
# Scale all retained genes and run PCA ------------------------------------
# Scaling all retained genes uses more memory than scaling variable genes only.
all.genes <- rownames(seurat_merged)
seurat_merged <- ScaleData(seurat_merged, features = all.genes)
seurat_merged <- RunPCA(
  seurat_merged, features = VariableFeatures(seurat_merged), seed.use = 42
)
# [["pca"]] accesses the stored PCA reduction.
print(seurat_merged[["pca"]], dims = 1:5, nfeatures = 5)
capture.output(print(seurat_merged[["pca"]], dims = 1:5, nfeatures = 5),
               file = file.path(output_dir, "pca_top_loadings.txt"))
pca_loadings <- VizDimLoadings(seurat_merged, dims = 1:2, reduction = "pca")
print(pca_loadings)
ggsave(file.path(output_dir, "pca_loadings.png"), pca_loadings,
       width = 10, height = 5, dpi = 300)
pca_plot <- DimPlot(seurat_merged, reduction = "pca", group.by = "sample_label")
print(pca_plot)
ggsave(file.path(output_dir, "pca_by_sample.png"), pca_plot,
       width = 7, height = 5, dpi = 300)
# fast=FALSE changes the drawing method so the same heatmap can be saved.
pca_heatmap <- DimHeatmap(
  seurat_merged, dims = 1, cells = 500, balanced = TRUE, fast = FALSE
)
print(pca_heatmap)
ggsave(file.path(output_dir, "pca_heatmap.png"), pca_heatmap,
       width = 10, height = 7, dpi = 300)

```

## Calculate UMAP and save the merged object

Calculate UMAP using PCs 1–10, retaining sample colours. Save the UMAP,
final cell metadata, session information and `seurat_merged_saved.rds`.
The final save is the stopping point. No clustering, marker detection,
cell-type annotation or integration follows it in this workflow.

```{r merged-save, eval=FALSE}
# UMAP and the final saved object -----------------------------------------
# PCs 1-10 are an initial setting to review for the supplied data.
seurat_merged <- RunUMAP(seurat_merged, dims = 1:10, seed.use = 42)
umap_plot <- DimPlot(seurat_merged, reduction = "umap", group.by = "sample_label")
print(umap_plot)
ggsave(file.path(output_dir, "umap_by_sample.png"), umap_plot,
       width = 7, height = 5, dpi = 300)
write.csv(seurat_merged[[]], file.path(output_dir, "cell_metadata_final.csv"))
capture.output(sessionInfo(), file = file.path(output_dir, "sessionInfo_final.txt"))
message("Saving the merged object. No analysis after STOP HERE is included.")
saveRDS(seurat_merged, file = file.path(output_dir, "seurat_merged_saved.rds"))
# STOP HERE
```

# Outputs and interpretation

The code above specifies outputs that will be created only when the relevant
analysis chunks are intentionally run. This document contains no executed
analysis results.

| Output | What to inspect after a run |
|---|---|
| `cell_counts_by_stage.csv` | Cells retained at each stage and, for the merge, in each sample. |
| Cell metadata CSV files | Per-cell counts, detected features and preserved sample identities. |
| QC and variable-feature PNG files | Quality distributions, count relationships and selected features. |
| PCA plots, heatmaps and loadings | Genes and cell patterns contributing to leading components. |
| UMAP PNG file | Exploratory cell relationships, interpreted with sample and quality information. |
| Session records and saved RDS | Software environment and the final object for later work. |

For a future run, record the actual retained counts, reasons for parameter
choices, observations from each plot and any unresolved limitations before
drawing conclusions. Keep interpretation separate from the method description.
`results/`, input data and local session files are ignored by Git. The generated
HTML guide is also ignored; the `.Rmd` is the shared source. Review generated
metadata separately before sharing it.

When changing analysis code, update its matching `.R` script and the
corresponding chunks here together so these two representations remain in sync.
